A Constraint Solving Approach to Parikh Images of Regular Languages

### 2022-07-07 main.tex
Evaluation

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{ebproof}

@@

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances generated
by the \Ostrich{} string constraint solver when solving the \Fudge{pyex
benchmarks}. \Fudge{The instances are all enormous and have some on average 900
products of 400000 automata with 97 counters each and 9999 constraints}. After
generating an initial~\InitialNrBenchmarks{}, we remove~\NrBroken{} misgenerated
instances that did not parse, as well as~\NrTrivial{} instances solved in under
five seconds by the baseline backend. The benchmarks are run on
commit~\texttt{\commit}.

The benchmarks are executed in parallel using GNU Parallel~\cite{parallel},
since they are mostly single-threaded and initial results showed negligible
interference on performance, on \BenchmarkRig{}. We compiled the code using
Scala~\ScalaVersion{}, and executed the experiments on~\JvmVersion{} with a
maximum heap of~\MaxHeapSize{}. We used \Nuxmv{} version~\NuxmvVersion{} invoked
as a subprocess for each instance. Experiments were executed on a fresh JVM per
backend, but the JVM was kept active for each batch of \BatchSize{} instances,
meaning that JIT compilation would make subsequent executions faster, while
still periodically restarting the JVM. We believe this represents a realistic
use case where \Calculus{} is integrated into a string solver or other system
and is called repeatedly. To mitigate systemic impacts of this decision on our
results, experiments were executed in random order for all backends. We run all
benchmarks with a timeout of~\RuntimeTimeout{}.

As a way of gauging the differences in overhead between the native backends and
\Nuxmv, we executed a small experiment where we ran the same trivial instance 30
times and observed the runtimes for all solvers. All backends saw an improvement
from JIT compilation within the first three runs, but subsequent runs only
improved the native backends. The final measured overhead for the fully warm
JVMs was less than~\OverheadSeconds. From this we draw the conclusion that major
differences in runtime between the native backends and \Nuxmv{} is more likely
to stem from innate differences rather than unfair overhead resulting from our
particular configuration. Additionally, one of the advantages of \Calculus{} to
\Nuxmv{} is that the former can me integrated into an automated theorem prover
while the latter has to be called as a separate process. In this sense, our
benchmarking setup can be said to be realistic.

All runtimes are measured in wall-clock time as observed by the JVM when executing the instance, and exclude time spent parsing, which in all observed cases was minimal in comparison to the runtime, usually far below 0.1 second.

It should be noted that despite running in the standard, deterministic
configuration, we observed some degree of nondeterminism in \Nuxmv. Instances
that were solved or unsolved would sometimes be solved or timed out in
subsequent runs. This observation was made even without randomised execution
order, suggesting that the nondeterminism likely stemmed either from our
encoding of the instances into \Nuxmv's format, or the execution environment.
Similarly, \Calculus{} showed similar signs of nondeterminism, but only for
satisfiable instances. To address this, we increased the runtime from our
initial experiments to give us a wider marigin. By contrast, the baseline
implementation showed no signs of nondeterminism. It should be noted that for
both backends exhibiting nondeterminism, the nondeterminism did not change the
general trend of the experiments, even for experiments with few instances.
Therefore, we believe that our large set of benchmarks will protect us from bias
since any backend is unlikely to be neither consistently lucky nor consistently
unlucky.

@@

In Figures~\ref{fig:solve-division} we show how many of the~\NrBenchmarks{} the
respective back-ends could solve and with which status. A full summary of their
outcomes is also available in Table~\ref{tab:solve-status}. We see here that
\Calculus{} generally outperforms \Nuxmv{} on unsatisfiable instances, while
being slightly worse at satisfiable instances. Both \Nuxmv{} and \Calculus{}
massively and strictly outperform the baseline approach on every kind of
instance, but most of all on satisfiable instances. In fact, the baseline
approach was significantly harder to benchmark due to its much higher demand for
RAM.

@@

% \begin{figure}[ht]
%   \begin{minipage}[b]{0.75\linewidth}
%     \centering 
%       \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-baseline-scatter}
%       \caption{Runtime duel: \Calculus{} versus baseline.}
%       \includegraphics[width=\textwidth]{graphs/\commit-duels-lazy-nuxmv-scatter}
%       \caption{Runtime duel: \Calculus{} versus \Nuxmv.}
%     \end{minipage}  
%   \caption{The pairwise runtime of each instance for instances solved by any backend. Timeouts are reported as a runtime of \RuntimeTimeout.}
%   \label{fig:duels}
% \end{figure}

@@

A cactus plot showing the number of instances solved within a given timeout for each backend can be seen in Figure~\ref{fig:cactus}. \Fudge{We see here that some solver is fast and some other is not}.

@@

The most obvious threats to validity would be poor benchmarking or poor
implementation, e.g. if the method described in \cref{sec:calculus,sec:multiple}
deviate from what is actually benchmarked in section \cref{sec:experiments}, or
if the methods used for benchmarking would be unsound. The first problem cannot
easily be addressed, except by stating that we have been careful in our
implementation and descriptions alike. The second is more practical. To increase
the probability that our results are representative both in the sense of
representing runs on expected, real world input and in the statistical sense for
a given run of \Catra{} despite the use of randomness in our
implementation we execute many experiments. Moreover, we have validated all
reported solutions made by \Calculus{} with \Nuxmv{} to ensure that \Calculus{}
is indeed sound and does not cheat.

@@

The most probable threat to validity is our choice of implementation platform
and our choice of automata library, that is our choice not to use one. Since our
automata are more advanced than anything supported by e.g. the BRICS library, we
opted to use our own implementation. There are some signs that in particular our
product computation is inefficient, notably the good performance of
low-threshold automata materialisation that prioritises computing smaller
products. This means that our implementation of \Calculus{} makes much less
heavy use of our product construction library than the baseline implementation
does, instead putting its weight into \Princess{}' theorem proving abilities.
This situation would unfairly advantage \Calculus{} if our automata library was
indeed the bottleneck. We believe this is unlikely since similar performance
issues as the ones in baseline in our string solver was the instigating reason
for this research. Additionally, while we may have a constant-factor performance
impact from bad implementation, it is less likely for our potentially poor
implementation to be asymptotically worse than e.g. BRICS. However, the
performance improvement of \Calculus{} over baseline seen in
\cref{sec:experiments} is generally not constant. Additionally, non-systematic
profiling suggests that both \Princess{}-based backends spend most of their time
in \Princess{}, suggesting that the automaton implementation is not the
bottleneck.

@@

In this paper we have introduced an automated theorem prover-friendly calculus
to compute commutating operations on intersections of regular languages that we
call \Calculus{}. We have evaluated it on \NrBenchmarks{} Parikh automata
intersection problems generated \Fudge{by the Ostrich+} string solver
\cite{ostrich} solving \Fudge{standard string constraint benchmarks} using our
Parikh automata solver \Catra{}.

>>>>

\usepackage{amsmath,empheq,fancybox}
\usepackage{paralist}
\usepackage{url}
\usepackage{color}
\usepackage{textcomp,listings}
\usepackage{array}
\usepackage{mymacros}
\usepackage{microtype}
\usepackage{listings}
\usepackage{csquotes}
\usepackage{proof}
\usepackage{cleveref}
\usepackage{algorithm2e}      
\usepackage{multirow}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{numprint}
\usepackage{ebproof}

@@

We evaluate the performance of \Catra{} on~\NrBenchmarks{} instances of Parikh
automata intersection problems generated by the \Ostrich{} string constraint
solver when solving the \Fudge{pyex benchmarks}. After generating an
initial~\InitialNrBenchmarks{}, we remove~\NrBroken{} misgenerated instances
that did not parse, as well as~\NrTrivial{} instances solved in under five
seconds by baseline. The benchmarks are run on commit~\texttt{\commit}.

The benchmarks are executed in parallel using GNU Parallel~\cite{parallel},
since they are mostly single-threaded and initial results showed negligible
interference on performance, on \BenchmarkRig{}. We compiled the code using
Scala~\ScalaVersion{}, and executed the experiments on~\JvmVersion{} with a
maximum heap of~\MaxHeapSize{}. We used \Nuxmv{} version~\NuxmvVersion{} invoked
as a subprocess for each instance. Instances were executed in batches of
\BatchSize{}, each given a fresh JVM. We believe this represents a realistic use
case where \Calculus{} is used to support e.g a string solver. Experiments were
executed in random order for all backends. Each instance got a time budget
of~\RuntimeTimeout.

All runtimes are measured in wall-clock time as observed by the JVM when
executing the instance, and exclude time spent parsing (usually far below
\numprint{0.1}s).

@@

In \cref{fig:solve-division} we show how many of the~\NrBenchmarks{} instances
the respective back-ends could solve and with which status. A full summary of
their outcomes is also available in \cref{tab:solve-status}. We see that
\Calculus{} generally outperforms \Nuxmv{} on unsatisfiable instances, while
being worse at satisfiable instances. Both \Nuxmv{} and \Calculus{} strictly
outperform baseline on every kind of instance, but most of all on satisfiable
instances. \Cref{tab:solve-status} also shows a small number of bugs in
\Calculus{}: 27~satisfiable instances are misreported as \Unsat{}, and 20
additional instances fail with an error. Both problems are due to
non-deterministic bugs in the clause learning. Note that \Calculus{} is sound:
it never misreports an unsatisfiable instance as \Sat{}, and all
\Fudge{\numprint{6218}} satisfying assignments have been verified by \Nuxmv{}.

@@

@@

A cactus plot showing the number of instances solved within a given timeout for
each backend can be seen in \cref{fig:cactus}. Here, we see clearly that
\Calculus{} outperforms \Nuxmv{} for shorter timeouts, and that \Nuxmv{} is
only able to catch up once we increase the timeout to far beyond what would be
normal for a string solver.

@@

The most obvious threats to validity would be poor benchmarking or poor
implementation, e.g. if the method described in \cref{sec:calculus,sec:multiple}
deviate from what is actually benchmarked in section \cref{sec:experiments}, or
if the methods used for benchmarking would be unsound.  To increase the
probability that our results are representative both in the sense of
representing  on expected inputs and in the statistical sense for a given run of
\Catra{} despite the use of randomness in our implementation we execute many
experiments. Moreover, to address the issue of correctness we have validated all
reported solutions made by \Calculus{} with \Nuxmv{} to ensure that \Calculus{}
is indeed sound.

@@

The greatest threat to validity is our choice of implementation platform and
automata library. There are some signs that product computation is inefficient,
notably the good performance of low-threshold automata materialisation that
prioritises computing smaller products. This means that \Calculus{} makes less
heavy use of product computation than baseline does, instead relying more on
\Princess{}'. This situation would unfairly advantage \Calculus{} if our
automata library was the bottleneck due to our automata library being
unnaturally poor. We believe this is unlikely since similar performance issues
have been reported for string solvers. Additionally, profiling suggests that
both \Princess{}-based backends spend most of their time in \Princess{},
suggesting that the automaton implementation is not the bottleneck.

@@

In this paper we have introduced a calculus to compute commutating operations on
intersections of regular languages that we call \Calculus{}. We have evaluated
it on \NrBenchmarks{} Parikh automata intersection problems generated \Fudge{by
the Ostrich+} string solver \cite{ostrich} solving \Fudge{standard string
constraint benchmarks} using our Parikh automata solver \Catra{}.
